---
title: "ST606 Project - Data Cleaning & Preparation"
author: "Student Name: Kessiya Mary Jose, Student ID: 25250360"
date: "`r Sys.Date()`"
output: html_document
---

## Project Topic: Do Weather Conditions Affect Wildlife Sightings in Ireland?

### Supervisor: Prof. Rafael de Andrade Moral


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE,
                      cache.path = "cache/")
```

```{r include=FALSE}
library(tidyverse)
```


#### Loading the Rainfall data
```{r message=FALSE, warning=FALSE}
rainfall <- read.csv("MTM01.20260317141732.csv")
head(rainfall)
```

#### Cleaning the Rainfall data (2016-2022)
```{r message=FALSE, warning=FALSE}
clean_rainfall <- rainfall %>%
  rename(
    statistic_type = Statistic.Label,
    month_code     = TLIST.M1.,
    station        = Meteorological.Weather.Station,
    value          = VALUE
  ) %>%
  filter(statistic_type == "Total Rainfall") %>%
  filter(station %in% c("Dublin airport", "Casement")) %>%
  mutate(
    year  = as.integer(substr(as.character(month_code), 1, 4)),
    month = as.integer(substr(as.character(month_code), 5, 6))
  ) %>%
  filter(year >= 2016 & year <= 2022) %>%
  mutate(station = case_when(
    station == "Dublin airport" ~ "Dublin Airport",
    station == "Casement"       ~ "Casement Aerodrome",
    TRUE ~ station
  )) %>%
  select(station, year, month, rainfall_mm = value) %>%
  filter(!is.na(rainfall_mm))

head(clean_rainfall)

```


#### Loading the Temperature data 
```{r message=FALSE, warning=FALSE}
temperature <- read.csv("MTM02.20260317141807.csv")
head(temperature)
```


#### Cleaning the Temperature data (2016-2022)
```{r message=FALSE, warning=FALSE}
clean_temperature <- temperature %>%
  rename(
    statistic_type = Statistic.Label,
    month_code     = TLIST.M1.,
    station        = Meteorological.Weather.Station,
    value          = VALUE
  ) %>%
  filter(statistic_type == "Mean Temperature") %>%
  filter(station %in% c("Dublin airport", "Casement")) %>%
  mutate(
    year  = as.integer(substr(as.character(month_code), 1, 4)),
    month = as.integer(substr(as.character(month_code), 5, 6))
  ) %>%
  filter(year >= 2016 & year <= 2022) %>%
  mutate(station = case_when(
    station == "Dublin airport" ~ "Dublin Airport",
    station == "Casement"       ~ "Casement Aerodrome",
    TRUE ~ station
  )) %>%
  select(station, year, month, mean_temp_celsius = value) %>%
  filter(!is.na(mean_temp_celsius))

head(clean_temperature)

```


#### Combining Rainfall and Temperature data (2016-2022)
```{r message=FALSE, warning=FALSE}
weather_2016_2022 <- clean_rainfall %>% 
inner_join(clean_temperature, by = c("station", "year", "month"))

head(weather_2016_2022)

```

#### Loading weather data (2023-2025)
```{r message=FALSE, warning=FALSE}
weather_2023_2025 <- read.csv("weather_2stations_2023_2025.csv")
head(weather_2023_2025)
```

##### Note: Since the weather data is unavilable in the file from years 2023 onwards, I took these data directly from the website for Dublin Port and Casement Aerodrome stations.


#### Cleaning weather data (2023-2025)
```{r message=FALSE, warning=FALSE}
weather_2023_2025 <- weather_2023_2025 %>%
  filter(year >= 2023 & year <= 2025) %>%
  select(station, year, month, rainfall_mm, mean_temp_celsius)

head(weather_2023_2025)

```


#### Combining weather data from 2016-2022 & 2023-2024 
```{r message=FALSE, warning=FALSE}
weather <- bind_rows(weather_2016_2022, weather_2023_2025) %>%
  arrange(station, year, month) %>% 
  filter(year <= 2023)

head(weather)
tail(weather)

```

##### Note: Since the Mammals data is unavilable from 2024 June onwards, I decided prepare the data from 2016 January to 2023 December.


#### Loading Mammals data
```{r message=FALSE, warning=FALSE}
mammals <- read.delim("MammalsOfIreland2016-2025.txt",
                           sep = "\t",
                           stringsAsFactors = FALSE)

head(mammals)

```


#### Cleaning Mammals data
```{r message=FALSE, warning=FALSE}
clean_mammals <- mammals %>%
  select(
    date      = Date,
    species   = TaxonName,
    county    = County,
    abundance = Abundance
  ) %>%
  mutate(date = as.Date(date, format = "%d/%m/%Y")) %>%
  mutate(
    year  = as.integer(format(date, "%Y")),
    month = as.integer(format(date, "%m"))
  ) %>%
  filter(year >= 2016 & year <= 2023) %>%
  filter(county %in% c("Dublin", "Kildare", "Wicklow",
                        "Meath", "Carlow", "Wexford")) %>%
  mutate(abundance = as.numeric(abundance)) %>%
  mutate(abundance = ifelse(is.na(abundance), 1, abundance)) %>%
  filter(county != "" & !is.na(county))

head(clean_mammals)

table(clean_mammals$county)

```


#### Monthly count of mammals
```{r message=FALSE, warning=FALSE}
mammals_monthly <- clean_mammals %>%
  group_by(year, month, county, species) %>%
  summarise(
    total_sightings = n(),
    total_abundance = sum(abundance, na.rm = TRUE),
    .groups = "drop"
  )

head(mammals_monthly)

```


#### Assigning weather stations to county
```{r message=FALSE, warning=FALSE}
mammals_monthly <- mammals_monthly %>%
  mutate(station = case_when(
    county %in% c("Dublin", "Wicklow", "Meath")   ~ "Dublin Airport",
    county %in% c("Kildare", "Carlow", "Wexford") ~ "Casement Aerodrome",
    TRUE ~ NA_character_
  ))

head(mammals_monthly)

```


#### Merging all data (Weather and Mammals)
```{r message=FALSE, warning=FALSE}
merged_dataset <- mammals_monthly %>%
  left_join(weather, by = c("station", "year", "month")) %>%
  filter(!is.na(rainfall_mm))

head(merged_dataset)

summary(merged_dataset)
```


#### Saving all files
```{r}
write.csv(weather, "weather_final_2016_2023.csv", row.names = FALSE)

write.csv(mammals_monthly, "mammals_monthly_summary.csv", row.names = FALSE)

write.csv(merged_dataset, "final_merged_dataset.csv",    row.names = FALSE)

```

