---
title: "ST606 Project - 5. JAGS Modelling"
author: "Kessiya Mary Jose - 25250360"
date: "`r format(Sys.time(), '%X %d %B, %Y')`"
output: html_document
---

#### Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE,
                      cache.path = "cache/")
library(tidyverse)
library(rjags)
library(coda)
library(parallel)

options(mc.cores = 6)
```


#### Loading dataset
```{r message=FALSE, include=FALSE}
final_data <- read.csv("final_dataset_expanded2.csv")

cat("Dataset Rows:", nrow(final_data), "\n")
cat("Columns:", names(final_data), "\n")
```


### Dataset preparation for fox, hare & marten species
```{r message=FALSE, warning=FALSE}
# converting sites to numeric site index
final_data <- final_data %>%
  dplyr::mutate(
    site      = as.integer(factor(station)),
    temp_scaled = as.numeric(scale(mean_temp))
  )

# checking the  number of sites
cat("Number of sites:", max(final_data$site), "\n")
cat("Site mapping:\n")
final_data %>%
  dplyr::select(station, site) %>%
  dplyr::distinct() %>%
  dplyr::arrange(site) %>%
  print()

# Creating 3 different dataset for each species
fox_data <- final_data %>%
  dplyr::filter(species == "Vulpes vulpes") %>%
  dplyr::filter(total_sightings >= 1)

hare_data <- final_data %>%
  dplyr::filter(species == "Lepus timidus subsp. hibernicus") %>%
  dplyr::filter(total_sightings >= 1)

marten_data <- final_data %>%
  dplyr::filter(species == "Martes martes") %>%
  dplyr::filter(total_sightings >= 1)

cat("Fox observations:", nrow(fox_data), "\n")
cat("Hare observations:", nrow(hare_data), "\n")
cat("Pine marten observations:", nrow(marten_data), "\n")
```


#### JAGS model code for fox (Vulpes vulpes)
```{r message=FALSE, warning=FALSE}
# The model has: - a truncated Poisson response (no zeros allowed) - a quadratic temperature effect (captures the parabola shape) - random intercept per site (each site has its own baseline) - random slope per site (each site responds differently to temperature) - random quadratic per site

# I am writing the model as a text string first and then JAGS will read it & fit it to the data

jags_model_code <- "
model {
  # Likelihood
  for (i in 1:N) {
    y[i] ~ dpois(mu[i]) T(1,)
    log(mu[i]) <- (beta0 + b0[site[i]]) +
                  (beta1 + b1[site[i]]) * temp[i] +
                  (beta2 + b2[site[i]]) * temp[i]^2
  }

  # Random effects for each site
  for (s in 1:S) {
    b0[s] ~ dnorm(0, tau0)
    b1[s] ~ dnorm(0, tau1)
    b2[s] ~ dnorm(0, tau2)
  }

  # Fixed effect priors (flat normal = very wide, letting data decide)
  beta0 ~ dnorm(0, 0.001)
  beta1 ~ dnorm(0, 0.001)
  beta2 ~ dnorm(0, 0.001)

  # Priors for variance of random effects (half-t approximated as gamma)
  tau0 ~ dgamma(0.1, 0.1)
  tau1 ~ dgamma(0.1, 0.1)
  tau2 ~ dgamma(0.1, 0.1)

  # Converting precision to standard deviation for reporting
  sigma0 <- 1 / sqrt(tau0)
  sigma1 <- 1 / sqrt(tau1)
  sigma2 <- 1 / sqrt(tau2)
}
"

# Saving the model code to a text file for JAGS to read it
writeLines(jags_model_code, "jags_model.txt")
```


#### Preparing fox data for JAGS
```{r message=FALSE, warning=FALSE}
# Since JAGS does not work with data frames it will need a list where each element is a vector or a number

fox_jags_data <- list(
  y    = fox_data$total_sightings,   # response variable
  temp = fox_data$temp_scaled,        # scaled temperature predictor
  site = fox_data$site,               # site index (1 to 9)
  N    = nrow(fox_data),              # total number of observations
  S    = max(fox_data$site)           # number of sites
)

cat("Fox data prepared for JAGS:\n")
cat("N (observations):", fox_jags_data$N, "\n")
cat("S (sites):", fox_jags_data$S, "\n")
cat("Min sightings:", min(fox_jags_data$y), "\n")
cat("Max sightings:", max(fox_jags_data$y), "\n")
```


#### Fitting the truncated Poisson model for fox in JAGS
```{r message=FALSE, warning=FALSE}
# Setting initial values to help JAGS start in a reasonable place
inits <- list(
  beta0 = 0,
  beta1 = 0,
  beta2 = 0,
  tau0  = 1,
  tau1  = 1,
  tau2  = 1
)

# Setting a random seed for reproducible results
set.seed(123)

# Compiling the model
fox_model <- jags.model(
  file   = "jags_model.txt",
  data   = fox_jags_data,
  inits  = inits,
  n.chains = 3,        # running 3 parallel chains for convergence checking
  n.adapt  = 1000      # adaptation phase
)

cat("Model compiled successfully\n")
```


#### Running the MCMC sampler
```{r message=FALSE, warning=FALSE}
# Burn-in phase
update(fox_model, n.iter = 2000)

# Sampling phase - keeping these samples for inference
fox_samples <- coda.samples(
  model          = fox_model,
  variable.names = c("beta0", "beta1", "beta2",
                     "sigma0", "sigma1", "sigma2"),
  n.iter         = 5000
)

cat("MCMC sampling complete\n")

# Checking convergence using R-hat (Gelman-Rubin statistic)
# R-hat close to 1.0 means the chains have converged
# R-hat above 1.1 means convergence problems
cat("\nConvergence check (R-hat values):\n")
print(gelman.diag(fox_samples))

# Summary of results
cat("\nModel summary:\n")
print(summary(fox_samples))
```


#### Saving fox model to csv file
```{r message=FALSE, warning=FALSE}
fox_summary <- summary(fox_samples)

# Extracting the mean estimates and credible intervals
fox_results <- data.frame(
  parameter = rownames(fox_summary$statistics),
  mean      = fox_summary$statistics[, "Mean"],
  sd        = fox_summary$statistics[, "SD"],
  lower_95  = fox_summary$quantiles[, "2.5%"],
  upper_95  = fox_summary$quantiles[, "97.5%"]
)

print(fox_results)
write.csv(fox_results, "fox_model_results.csv")
```


#### Plotting the predicted temperature effect for fox
```{r message=FALSE, warning=FALSE}
# I am creating a sequence of temperature values and then calculating the predicted mean sightings at each temperature

# Getting the posterior mean estimates
beta0_mean <- fox_results$mean[fox_results$parameter == "beta0"]
beta1_mean <- fox_results$mean[fox_results$parameter == "beta1"]
beta2_mean <- fox_results$mean[fox_results$parameter == "beta2"]

# Creating a sequence of scaled temperatures to predict over
temp_seq <- seq(min(fox_data$temp_scaled),
                max(fox_data$temp_scaled),
                length.out = 100)

# Calculating predicted mean sightings on original scale
predicted_mu <- exp(beta0_mean +
                    beta1_mean * temp_seq +
                    beta2_mean * temp_seq^2)

# Converting scaled temperature back to Celsius for the x axis
temp_mean_val <- mean(fox_data$mean_temp)
temp_sd_val   <- sd(fox_data$mean_temp)
temp_celsius  <- temp_seq * temp_sd_val + temp_mean_val

# Creating the plot
pred_df <- data.frame(
  temperature = temp_celsius,
  predicted   = predicted_mu
)

ggplot(pred_df, aes(x = temperature, y = predicted)) +
  geom_line(colour = "#E53935", linewidth = 1.2) +
  geom_rug(data = fox_data,
           aes(x = mean_temp, y = NULL),
           alpha = 0.3) +
  labs(
    title    = "Fox sightings - predicted temperature effect",
    subtitle = "Truncated Poisson Bayesian model, posterior mean",
    x        = "Mean monthly temperature (°C)",
    y        = "Predicted mean sightings"
  ) +
  theme_minimal()

ggsave("plot_fox_temperature_effect.png", width = 8, height = 5)
```


#### Fitting the truncated Poisson model for hare
```{r message=FALSE, warning=FALSE}
hare_jags_data <- list(
  y    = hare_data$total_sightings,
  temp = hare_data$temp_scaled,
  site = hare_data$site,
  N    = nrow(hare_data),
  S    = max(hare_data$site)
)

cat("Hare data prepared for JAGS:\n")
cat("N (observations):", hare_jags_data$N, "\n")
cat("S (sites):", hare_jags_data$S, "\n")
cat("Min sightings:", min(hare_jags_data$y), "\n")
cat("Max sightings:", max(hare_jags_data$y), "\n")

set.seed(123)

hare_model <- jags.model(
  file     = "jags_model.txt",
  data     = hare_jags_data,
  inits    = inits,
  n.chains = 3,
  n.adapt  = 1000
)

cat("Hare model compiled\n")
update(hare_model, n.iter = 2000)

hare_samples <- coda.samples(
  model          = hare_model,
  variable.names = c("beta0", "beta1", "beta2",
                     "sigma0", "sigma1", "sigma2"),
  n.iter         = 5000
)

cat("Hare MCMC complete\n")
cat("\nHare convergence check:\n")
print(gelman.diag(hare_samples))
cat("\nHare model summary:\n")
print(summary(hare_samples))
```


#### Fitting the truncated Poisson model for pine marten
```{r message=FALSE, warning=FALSE}
marten_jags_data <- list(
  y    = marten_data$total_sightings,
  temp = marten_data$temp_scaled,
  site = marten_data$site,
  N    = nrow(marten_data),
  S    = max(marten_data$site)
)

cat("Pine marten data prepared for JAGS:\n")
cat("N (observations):", marten_jags_data$N, "\n")
cat("S (sites):", marten_jags_data$S, "\n")
cat("Min sightings:", min(marten_jags_data$y), "\n")
cat("Max sightings:", max(marten_jags_data$y), "\n")

set.seed(123)

marten_model <- jags.model(
  file     = "jags_model.txt",
  data     = marten_jags_data,
  inits    = inits,
  n.chains = 3,
  n.adapt  = 1000
)

cat("Pine marten model compiled.\n")
update(marten_model, n.iter = 2000)

marten_samples <- coda.samples(
  model          = marten_model,
  variable.names = c("beta0", "beta1", "beta2",
                     "sigma0", "sigma1", "sigma2"),
  n.iter         = 5000
)

cat("Pine marten MCMC complete.\n")
cat("\nPine marten convergence check:\n")
print(gelman.diag(marten_samples))
cat("\nPine marten model summary:\n")
print(summary(marten_samples))
```


#### Fixing hare convergence
```{r message=FALSE, warning=FALSE}
# The hare model needs more iterations because beta0 R-hat is 1.24. So I am running 5000 more burn-in iterations and 10000 sampling iterations to help the chains converge properly

update(hare_model, n.iter = 5000)

hare_samples2 <- coda.samples(
  model          = hare_model,
  variable.names = c("beta0", "beta1", "beta2",
                     "sigma0", "sigma1", "sigma2"),
  n.iter         = 10000
)

cat("Extended hare MCMC complete!\n")
cat("\nHare convergence check after extended run:\n")
print(gelman.diag(hare_samples2))
cat("\nHare model summary after extended run:\n")
print(summary(hare_samples2))
```


#### Saving hare and martine's files
```{r message=FALSE, warning=FALSE}
# Hare - using the extended samples (hare_samples2)
hare_summary <- summary(hare_samples2)
hare_results <- data.frame(
  parameter = rownames(hare_summary$statistics),
  mean      = hare_summary$statistics[, "Mean"],
  sd        = hare_summary$statistics[, "SD"],
  lower_95  = hare_summary$quantiles[, "2.5%"],
  upper_95  = hare_summary$quantiles[, "97.5%"]
)
write.csv(hare_results, "hare_model_results.csv")

# Pine marten
marten_summary <- summary(marten_samples)
marten_results <- data.frame(
  parameter = rownames(marten_summary$statistics),
  mean      = marten_summary$statistics[, "Mean"],
  sd        = marten_summary$statistics[, "SD"],
  lower_95  = marten_summary$quantiles[, "2.5%"],
  upper_95  = marten_summary$quantiles[, "97.5%"]
)
write.csv(marten_results, "marten_model_results.csv")

print(hare_results)
print(marten_results)
```


#### Combined prediction plot of three species with temperature effect
```{r message=FALSE, warning=FALSE}
# Function to calculate predictions for a species
get_predictions <- function(results, species_data, species_name) {
  b0 <- results$mean[results$parameter == "beta0"]
  b1 <- results$mean[results$parameter == "beta1"]
  b2 <- results$mean[results$parameter == "beta2"]

  temp_seq <- seq(min(species_data$temp_scaled),
                  max(species_data$temp_scaled),
                  length.out = 100)

  temp_mean_val <- mean(species_data$mean_temp)
  temp_sd_val   <- sd(species_data$mean_temp)
  temp_celsius  <- temp_seq * temp_sd_val + temp_mean_val

  predicted <- exp(b0 + b1 * temp_seq + b2 * temp_seq^2)

  data.frame(
    temperature = temp_celsius,
    predicted   = predicted,
    species     = species_name
  )
}

# Getting predictions for each species
fox_pred    <- get_predictions(fox_results,    fox_data,    "Vulpes vulpes (Fox)")
hare_pred   <- get_predictions(hare_results,   hare_data,   "Lepus timidus (Hare)")
marten_pred <- get_predictions(marten_results, marten_data, "Martes martes (Pine marten)")

# Combining into one data frame
all_preds <- dplyr::bind_rows(fox_pred, hare_pred, marten_pred)

# Plotting all three species together
ggplot(all_preds, aes(x = temperature, y = predicted, colour = species)) +
  geom_line(linewidth = 1.2) +
  facet_wrap(~species, scales = "free_y", ncol = 3) +
  labs(
    title    = "Predicted temperature effect on sightings - all three species",
    subtitle = "Truncated Poisson Bayesian mixed model, posterior mean",
    x        = "Mean monthly temperature (°C)",
    y        = "Predicted mean sightings"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

ggsave("plot_all_species_temperature.png", width = 14, height = 5)
```


#### Credible interval bands
```{r message=FALSE, warning=FALSE}
# By adding credible interval bands to the prediction plot instead of just the posterior mean line, i can show the 95% credible interval as a shaded band around each curve. This shows how uncertain the predictions are.

# Function to get predictions with credible intervals using all MCMC samples rather than just the mean
get_predictions_ci <- function(samples, species_data, species_name) {

  # Converting samples to a matrix
  samples_matrix <- as.matrix(samples)

  # Getting all posterior samples for beta0, beta1, beta2
  b0_samples <- samples_matrix[, "beta0"]
  b1_samples <- samples_matrix[, "beta1"]
  b2_samples <- samples_matrix[, "beta2"]

  # Temperature sequence
  temp_seq <- seq(min(species_data$temp_scaled),
                  max(species_data$temp_scaled),
                  length.out = 100)

  # Converting back to Celsius
  temp_mean_val <- mean(species_data$mean_temp)
  temp_sd_val   <- sd(species_data$mean_temp)
  temp_celsius  <- temp_seq * temp_sd_val + temp_mean_val

  # Calculating predicted values for every MCMC sample, which gives a distribution of predictions at each temperature
  pred_matrix <- sapply(temp_seq, function(t) {
    exp(b0_samples + b1_samples * t + b2_samples * t^2)
  })

  # Summarising across all samples
  data.frame(
    temperature = temp_celsius,
    predicted   = apply(pred_matrix, 2, median),
    lower_95    = apply(pred_matrix, 2, quantile, probs = 0.025),
    upper_95    = apply(pred_matrix, 2, quantile, probs = 0.975),
    species     = species_name
  )
}

# Getting predictions with credible intervals for each species
fox_pred_ci    <- get_predictions_ci(fox_samples,    fox_data,
                                     "Vulpes vulpes (Fox)")
hare_pred_ci   <- get_predictions_ci(hare_samples2,  hare_data,
                                     "Lepus timidus (Hare)")
marten_pred_ci <- get_predictions_ci(marten_samples, marten_data,
                                     "Martes martes (Pine marten)")

# Combining all three
all_preds_ci <- dplyr::bind_rows(fox_pred_ci, hare_pred_ci, marten_pred_ci)

# Plotting with credible interval bands
ggplot(all_preds_ci, aes(x = temperature)) +
  geom_ribbon(aes(ymin = lower_95, ymax = upper_95, fill = species),
              alpha = 0.2) +
  geom_line(aes(y = predicted, colour = species), linewidth = 1.2) +
  facet_wrap(~species, scales = "free_y", ncol = 3) +
  labs(
    title    = "Predicted temperature effect on sightings with 95% credible intervals",
    subtitle = "Truncated Poisson Bayesian mixed model",
    x        = "Mean monthly temperature (°C)",
    y        = "Predicted mean sightings"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

ggsave("plot_all_species_temperature_ci.png", width = 14, height = 5)
```


#### Checking convergence of fox
```{r message=FALSE, warning=FALSE, fig.width=12, fig.height=8}
# Trace plots show how the MCMC chains moved through the parameter space. 
# If the chains look like a "hairy caterpillar" mixing well, then it means good convergence
# If chains are drifting or stuck that means poor convergence

# Fox trace plots
plot(fox_samples[, c("beta0", "beta1", "beta2",
                     "sigma0", "sigma1", "sigma2")])
```


#Saving fox trace plots
```{r message=FALSE, warning=FALSE, fig.width=12, fig.height=8}
png("plot_fox_traces.png", width = 1400, height = 1800)
par(mfrow = c(6, 2))
plot(fox_samples[, c("beta0", "beta1", "beta2",
                     "sigma0", "sigma1", "sigma2")])
dev.off()
```


#### Checking convergence of hare
```{r message=FALSE, warning=FALSE, fig.width=12, fig.height=8}
# Hare trace plots
plot(hare_samples2[, c("beta0", "beta1", "beta2",
                       "sigma0", "sigma1", "sigma2")])
```


#Saving hare trace plots
```{r message=FALSE, warning=FALSE, fig.width=12, fig.height=8}
png("plot_hare_traces.png", width = 1400, height = 1800)
par(mfrow = c(6, 2))
plot(hare_samples2[, c("beta0", "beta1", "beta2",
                       "sigma0", "sigma1", "sigma2")])
dev.off()
```


#### Checking convergence of marten
```{r message=FALSE, warning=FALSE,  fig.width=12, fig.height=8}
# Pine marten trace plots
plot(marten_samples[, c("beta0", "beta1", "beta2",
                        "sigma0", "sigma1", "sigma2")])
```


#### Saving marten trace plots
```{r message=FALSE, warning=FALSE, fig.width=12, fig.height=8}
# Saving pine marten traces to file
png("plot_marten_traces.png", width = 1400, height = 1800)
par(mfrow = c(6, 2))
plot(marten_samples[, c("beta0", "beta1", "beta2",
                        "sigma0", "sigma1", "sigma2")])
dev.off()
```


#### Creating datasets with and without 2019
```{r message=FALSE, warning=FALSE}
# Creating datasets without 2019
fox_data_no2019 <- fox_data %>%
  dplyr::filter(year != 2019)

hare_data_no2019 <- hare_data %>%
  dplyr::filter(year != 2019)

marten_data_no2019 <- marten_data %>%
  dplyr::filter(year != 2019)

cat("Fox observations with 2019:", nrow(fox_data), "\n")
cat("Fox observations without 2019:", nrow(fox_data_no2019), "\n")
cat("\nHare observations with 2019:", nrow(hare_data), "\n")
cat("Hare observations without 2019:", nrow(hare_data_no2019), "\n")
cat("\nMarten observations with 2019:", nrow(marten_data), "\n")
cat("Marten observations without 2019:", nrow(marten_data_no2019), "\n")
```


Key findings
Fox: 0 sightings recorded in 2019
Hare: 14
Martin: 12



#### Refitting hare model without 2019
```{r message=FALSE, warning=FALSE}
hare_jags_data_no2019 <- list(
  y    = hare_data_no2019$total_sightings,
  temp = hare_data_no2019$temp_scaled,
  site = hare_data_no2019$site,
  N    = nrow(hare_data_no2019),
  S    = max(hare_data_no2019$site)
)

cat("Hare data without 2019 prepared:\n")
cat("N:", hare_jags_data_no2019$N, "\n")

set.seed(123)

hare_model_no2019 <- jags.model(
  file     = "jags_model.txt",
  data     = hare_jags_data_no2019,
  inits    = inits,
  n.chains = 3,
  n.adapt  = 1000
)

update(hare_model_no2019, n.iter = 5000)

hare_samples_no2019 <- coda.samples(
  model          = hare_model_no2019,
  variable.names = c("beta0", "beta1", "beta2",
                     "sigma0", "sigma1", "sigma2"),
  n.iter         = 10000
)

cat("Hare model without 2019 complete!\n")
cat("\nConvergence check:\n")
print(gelman.diag(hare_samples_no2019))
cat("\nResults:\n")
print(summary(hare_samples_no2019))
```


#### Refitting pine marten model without 2019 
```{r message=FALSE, warning=FALSE}
marten_jags_data_no2019 <- list(
  y    = marten_data_no2019$total_sightings,
  temp = marten_data_no2019$temp_scaled,
  site = marten_data_no2019$site,
  N    = nrow(marten_data_no2019),
  S    = max(marten_data_no2019$site)
)

cat("Pine marten data without 2019 prepared:\n")
cat("N:", marten_jags_data_no2019$N, "\n")

set.seed(123)

marten_model_no2019 <- jags.model(
  file     = "jags_model.txt",
  data     = marten_jags_data_no2019,
  inits    = inits,
  n.chains = 3,
  n.adapt  = 1000
)

update(marten_model_no2019, n.iter = 5000)

marten_samples_no2019 <- coda.samples(
  model          = marten_model_no2019,
  variable.names = c("beta0", "beta1", "beta2",
                     "sigma0", "sigma1", "sigma2"),
  n.iter         = 10000
)

cat("Pine marten model without 2019 complete!\n")
cat("\nConvergence check:\n")
print(gelman.diag(marten_samples_no2019))
cat("\nResults:\n")
print(summary(marten_samples_no2019))
```


#### Comparing results with and without 2019
```{r message=FALSE, warning=FALSE}
# If the estimates are similar, the 2019 gap does not affect conclusions

# Creating a comparison table
comparison <- data.frame(
  parameter = c("beta0", "beta1", "beta2"),
  
  # Hare with 2019
  hare_with_mean    = c(1.1478, -0.2851, -0.3468),
  hare_with_lower   = c(0.7638, -0.5900, -0.5718),
  hare_with_upper   = c(1.5686, 0.0243, -0.1250),
  
  # Hare without 2019
  hare_no2019_mean  = c(1.1492, -0.2875, -0.3943),
  hare_no2019_lower = c(0.8275, -0.6100, -0.6435),
  hare_no2019_upper = c(1.4643, -0.0217, -0.1549),
  
  # Marten with 2019
  marten_with_mean    = c(0.7061, 0.4107, -0.2931),
  marten_with_lower   = c(0.4295, 0.2154, -0.5928),
  marten_with_upper   = c(0.9722, 0.6148, -0.0491),
  
  # Marten without 2019
  marten_no2019_mean  = c(0.7235, 0.4147, -0.2782),
  marten_no2019_lower = c(0.4139, 0.1980, -0.5739),
  marten_no2019_upper = c(1.0066, 0.6426, -0.0120)
)

print(comparison)
write.csv(comparison, "comparison_with_without_2019.csv")
```


#### Visual comparison plot
```{r message=FALSE, warning=FALSE, fig.width=14, fig.height=6}
# Hare predictions without 2019
hare_no2019_summary <- summary(hare_samples_no2019)
hare_no2019_results <- data.frame(
  parameter = rownames(hare_no2019_summary$statistics),
  mean      = hare_no2019_summary$statistics[, "Mean"],
  sd        = hare_no2019_summary$statistics[, "SD"],
  lower_95  = hare_no2019_summary$quantiles[, "2.5%"],
  upper_95  = hare_no2019_summary$quantiles[, "97.5%"]
)

# Marten predictions without 2019
marten_no2019_summary <- summary(marten_samples_no2019)
marten_no2019_results <- data.frame(
  parameter = rownames(marten_no2019_summary$statistics),
  mean      = marten_no2019_summary$statistics[, "Mean"],
  sd        = marten_no2019_summary$statistics[, "SD"],
  lower_95  = marten_no2019_summary$quantiles[, "2.5%"],
  upper_95  = marten_no2019_summary$quantiles[, "97.5%"]
)

# Getting predictions for both versions
hare_pred_all    <- get_predictions(hare_results,
                                    hare_data,
                                    "Hare - all data")
hare_pred_no2019 <- get_predictions(hare_no2019_results,
                                    hare_data_no2019,
                                    "Hare - excluding 2019")

marten_pred_all    <- get_predictions(marten_results,
                                      marten_data,
                                      "Marten - all data")
marten_pred_no2019 <- get_predictions(marten_no2019_results,
                                      marten_data_no2019,
                                      "Marten - excluding 2019")

# Combining
comparison_preds <- dplyr::bind_rows(
  hare_pred_all, hare_pred_no2019,
  marten_pred_all, marten_pred_no2019
) %>%
  dplyr::mutate(
    animal   = ifelse(grepl("Hare", species), "Hare", "Pine marten"),
    dataset  = ifelse(grepl("all data", species),
                      "All data (including 2019)",
                      "Excluding 2019")
  )

# Plotting comparison
ggplot(comparison_preds,
       aes(x = temperature, y = predicted,
           colour = dataset, linetype = dataset)) +
  geom_line(linewidth = 1.1) +
  facet_wrap(~animal, scales = "free_y") +
  scale_colour_manual(values = c("#E53935", "#1565C0")) +
  labs(
    title    = "Comparison of predictions with and without 2019 data",
    subtitle = "Truncated Poisson Bayesian mixed model",
    x        = "Mean monthly temperature (°C)",
    y        = "Predicted mean sightings",
    colour   = "Dataset",
    linetype = "Dataset"
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")

ggsave("plot_comparison_2019.png", width = 12, height = 5)
```


#### Adding scaled rainfall to the species datasets
```{r message=FALSE, warning=FALSE}
final_data <- final_data %>%
  dplyr::mutate(rainfall_scaled = as.numeric(scale(rainfall_mm)))

# Recreating the three species datasets with the new scaled rainfall column
fox_data <- final_data %>%
  dplyr::filter(species == "Vulpes vulpes") %>%
  dplyr::filter(total_sightings >= 1)

hare_data <- final_data %>%
  dplyr::filter(species == "Lepus timidus subsp. hibernicus") %>%
  dplyr::filter(total_sightings >= 1)

marten_data <- final_data %>%
  dplyr::filter(species == "Martes martes") %>%
  dplyr::filter(total_sightings >= 1)

cat("Rainfall scaling is done.\n")
cat("Mean scaled rainfall:", round(mean(final_data$rainfall_scaled), 4), "\n")
cat("SD scaled rainfall:", round(sd(final_data$rainfall_scaled), 4), "\n")
cat("Fox rows:", nrow(fox_data), "\n")
cat("Hare rows:", nrow(hare_data), "\n")
cat("Marten rows:", nrow(marten_data), "\n")
```


#### New JAGS model that includes both temperature and rainfall
```{r message=FALSE, warning=FALSE}
# Rainfall is added as a fixed effect only (no random slope per site)
# beta3 is the rainfall effect

jags_model_rainfall <- "
model {
  # Likelihood
  for (i in 1:N) {
    y[i] ~ dpois(mu[i]) T(1,)
    log(mu[i]) <- (beta0 + b0[site[i]]) +
                  (beta1 + b1[site[i]]) * temp[i] +
                  (beta2 + b2[site[i]]) * temp[i]^2 +
                  beta3 * rainfall[i]
  }

  # Random effects for each site
  for (s in 1:S) {
    b0[s] ~ dnorm(0, tau0)
    b1[s] ~ dnorm(0, tau1)
    b2[s] ~ dnorm(0, tau2)
  }

  # Fixed effect priors
  beta0 ~ dnorm(0, 0.001)
  beta1 ~ dnorm(0, 0.001)
  beta2 ~ dnorm(0, 0.001)
  beta3 ~ dnorm(0, 0.001)

  # Priors for variance of random effects
  tau0 ~ dgamma(0.1, 0.1)
  tau1 ~ dgamma(0.1, 0.1)
  tau2 ~ dgamma(0.1, 0.1)

  # Converting precision to standard deviation
  sigma0 <- 1 / sqrt(tau0)
  sigma1 <- 1 / sqrt(tau1)
  sigma2 <- 1 / sqrt(tau2)
}
"

writeLines(jags_model_rainfall, "jags_model_rainfall.txt")
cat("Rainfall model code is saved as jags_model_rainfall.txt\n")
```


#### Fitting the rainfall model for fox
```{r message=FALSE, warning=FALSE}
fox_jags_data_rain <- list(
  y        = fox_data$total_sightings,
  temp     = fox_data$temp_scaled,
  rainfall = fox_data$rainfall_scaled,
  site     = fox_data$site,
  N        = nrow(fox_data),
  S        = max(fox_data$site)
)

inits_rain <- list(
  beta0 = 0, beta1 = 0,
  beta2 = 0, beta3 = 0,
  tau0  = 1, tau1  = 1, tau2 = 1
)

set.seed(123)

fox_model_rain <- jags.model(
  file     = "jags_model_rainfall.txt",
  data     = fox_jags_data_rain,
  inits    = inits_rain,
  n.chains = 3,
  n.adapt  = 1000
)

update(fox_model_rain, n.iter = 2000)

fox_samples_rain <- coda.samples(
  model          = fox_model_rain,
  variable.names = c("beta0", "beta1", "beta2", "beta3",
                     "sigma0", "sigma1", "sigma2"),
  n.iter         = 5000
)

cat("Fox rainfall model is complete\n")
cat("\nConvergence check:\n")
print(gelman.diag(fox_samples_rain))
cat("\nResults:\n")
print(summary(fox_samples_rain))
```


#### Fitting the rainfall model for hare

```{r message=FALSE, warning=FALSE}
hare_jags_data_rain <- list(
  y        = hare_data$total_sightings,
  temp     = hare_data$temp_scaled,
  rainfall = hare_data$rainfall_scaled,
  site     = hare_data$site,
  N        = nrow(hare_data),
  S        = max(hare_data$site)
)

set.seed(123)

hare_model_rain <- jags.model(
  file     = "jags_model_rainfall.txt",
  data     = hare_jags_data_rain,
  inits    = inits_rain,
  n.chains = 3,
  n.adapt  = 1000
)

update(hare_model_rain, n.iter = 5000)

hare_samples_rain <- coda.samples(
  model          = hare_model_rain,
  variable.names = c("beta0", "beta1", "beta2", "beta3",
                     "sigma0", "sigma1", "sigma2"),
  n.iter         = 10000
)

cat("Hare rainfall model is completed\n")
cat("\nConvergence check:\n")
print(gelman.diag(hare_samples_rain))
cat("\nResults:\n")
print(summary(hare_samples_rain))
```


#### Fitting the rainfall model for pine marten
```{r message=FALSE, warning=FALSE}
marten_jags_data_rain <- list(
  y        = marten_data$total_sightings,
  temp     = marten_data$temp_scaled,
  rainfall = marten_data$rainfall_scaled,
  site     = marten_data$site,
  N        = nrow(marten_data),
  S        = max(marten_data$site)
)

set.seed(123)

marten_model_rain <- jags.model(
  file     = "jags_model_rainfall.txt",
  data     = marten_jags_data_rain,
  inits    = inits_rain,
  n.chains = 3,
  n.adapt  = 1000
)

update(marten_model_rain, n.iter = 2000)

marten_samples_rain <- coda.samples(
  model          = marten_model_rain,
  variable.names = c("beta0", "beta1", "beta2", "beta3",
                     "sigma0", "sigma1", "sigma2"),
  n.iter         = 5000
)

cat("Pine marten rainfall model is completed!\n")
cat("\nConvergence check:\n")
print(gelman.diag(marten_samples_rain))
cat("\nResults:\n")
print(summary(marten_samples_rain))
```


#### Fox rainfall model - 2
```{r message=FALSE, warning=FALSE}
# Fox rainfall model needs more iterations because beta0 R-hat is 1.12 which is slightly above the 1.1 threshold

cat("Running extended MCMC for fox rainfall model...\n")

update(fox_model_rain, n.iter = 5000)

fox_samples_rain2 <- coda.samples(
  model          = fox_model_rain,
  variable.names = c("beta0", "beta1", "beta2", "beta3",
                     "sigma0", "sigma1", "sigma2"),
  n.iter         = 10000
)

cat("Extended fox rainfall MCMC is completed\n")
cat("\nConvergence check after extended run:\n")
print(gelman.diag(fox_samples_rain2))
cat("\nResults:\n")
print(summary(fox_samples_rain2))
```


#### Saving rainfall model results for all three species
```{r message=FALSE, warning=FALSE}
fox_rain_summary <- summary(fox_samples_rain2)
fox_rain_results <- data.frame(
  parameter = rownames(fox_rain_summary$statistics),
  mean      = fox_rain_summary$statistics[, "Mean"],
  sd        = fox_rain_summary$statistics[, "SD"],
  lower_95  = fox_rain_summary$quantiles[, "2.5%"],
  upper_95  = fox_rain_summary$quantiles[, "97.5%"]
)
write.csv(fox_rain_results, "fox_rainfall_model_results.csv")

hare_rain_summary <- summary(hare_samples_rain)
hare_rain_results <- data.frame(
  parameter = rownames(hare_rain_summary$statistics),
  mean      = hare_rain_summary$statistics[, "Mean"],
  sd        = hare_rain_summary$statistics[, "SD"],
  lower_95  = hare_rain_summary$quantiles[, "2.5%"],
  upper_95  = hare_rain_summary$quantiles[, "97.5%"]
)
write.csv(hare_rain_results, "hare_rainfall_model_results.csv")

marten_rain_summary <- summary(marten_samples_rain)
marten_rain_results <- data.frame(
  parameter = rownames(marten_rain_summary$statistics),
  mean      = marten_rain_summary$statistics[, "Mean"],
  sd        = marten_rain_summary$statistics[, "SD"],
  lower_95  = marten_rain_summary$quantiles[, "2.5%"],
  upper_95  = marten_rain_summary$quantiles[, "97.5%"]
)
write.csv(marten_rain_results, "marten_rainfall_model_results.csv")

# Printing beta3 (rainfall effect) for each species
cat("\n----- Rainfall effect (beta3) summary -----\n")
cat("\nFox beta3:\n")
print(fox_rain_results[fox_rain_results$parameter == "beta3", ])
cat("\nHare beta3:\n")
print(hare_rain_results[hare_rain_results$parameter == "beta3", ])
cat("\nMarten beta3:\n")
print(marten_rain_results[marten_rain_results$parameter == "beta3", ])
```


#### Rainfall effect plot for all three species
```{r message=FALSE, warning=FALSE, fig.width=14, fig.height=5}
# Creating a rainfall plot to show how sightings change with rainfall by holding temperature at its mean (scaled value = 0)

# Function to get rainfall predictions
get_rainfall_predictions <- function(results, species_data, species_name) {
  b0 <- results$mean[results$parameter == "beta0"]
  b1 <- results$mean[results$parameter == "beta1"]
  b2 <- results$mean[results$parameter == "beta2"]
  b3 <- results$mean[results$parameter == "beta3"]

  # rainfall sequence
  rain_seq <- seq(min(species_data$rainfall_scaled),
                  max(species_data$rainfall_scaled),
                  length.out = 100)

  # converting back to mm
  rain_mean_val <- mean(species_data$rainfall_mm)
  rain_sd_val   <- sd(species_data$rainfall_mm)
  rain_mm       <- rain_seq * rain_sd_val + rain_mean_val

  # predicting at mean temperature (scaled = 0)
  predicted <- exp(b0 + b1 * 0 + b2 * 0 + b3 * rain_seq)

  data.frame(
    rainfall  = rain_mm,
    predicted = predicted,
    species   = species_name
  )
}

fox_rain_pred    <- get_rainfall_predictions(
  fox_rain_results, fox_data, "Vulpes vulpes (Fox)")
hare_rain_pred   <- get_rainfall_predictions(
  hare_rain_results, hare_data, "Lepus timidus (Hare)")
marten_rain_pred <- get_rainfall_predictions(
  marten_rain_results, marten_data, "Martes martes (Pine marten)")

all_rain_preds <- dplyr::bind_rows(
  fox_rain_pred, hare_rain_pred, marten_rain_pred)

ggplot(all_rain_preds, aes(x = rainfall, y = predicted,
                            colour = species)) +
  geom_line(linewidth = 1.2) +
  facet_wrap(~species, scales = "free_y", ncol = 3) +
  labs(
    title    = "Predicted rainfall effect on sightings - all three species",
    subtitle = "Temperature held at mean value",
    x        = "Monthly rainfall (mm)",
    y        = "Predicted mean sightings"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

ggsave("plot_all_species_rainfall.png", width = 14, height = 5)
```


------------------------------------------------------------------------

#### Seasonal effects - Adding month as a factor variable to the species datasets
```{r message=FALSE, warning=FALSE}
# I am using month as an integer (1-12) which JAGS will treat as a category January (month 1) will be the baseline reference month.
# all other months get an effect relative to January

# Checking the month distribution in each species dataset
cat("Month distribution in fox data:\n")
print(table(fox_data$month))

cat("\nMonth distribution in hare data:\n")
print(table(hare_data$month))

cat("\nMonth distribution in marten data:\n")
print(table(marten_data$month))
```


#### JAGS model with temperature, rainfall and seasonal effects
```{r message=FALSE, warning=FALSE}
# Month is included as a categorical predictor
# alpha[month] gives a separate effect for each of the 12 months
# alpha[1] (January) is set to zero as the reference or baseline
# all other months are estimated relative to January

jags_model_seasonal <- "
model {
  # Likelihood
  for (i in 1:N) {
    y[i] ~ dpois(mu[i]) T(1,)
    log(mu[i]) <- (beta0 + b0[site[i]]) +
                  (beta1 + b1[site[i]]) * temp[i] +
                  (beta2 + b2[site[i]]) * temp[i]^2 +
                  beta3 * rainfall[i] +
                  alpha[month[i]]
  }

  # Random effects for each site
  for (s in 1:S) {
    b0[s] ~ dnorm(0, tau0)
    b1[s] ~ dnorm(0, tau1)
    b2[s] ~ dnorm(0, tau2)
  }

  # Monthly seasonal effects
  # January (month 1) is fixed at 0 as the reference month
  alpha[1] <- 0
  for (m in 2:12) {
    alpha[m] ~ dnorm(0, 0.001)
  }

  # Fixed effect priors
  beta0 ~ dnorm(0, 0.001)
  beta1 ~ dnorm(0, 0.001)
  beta2 ~ dnorm(0, 0.001)
  beta3 ~ dnorm(0, 0.001)

  # Priors for variance of random effects
  tau0 ~ dgamma(0.1, 0.1)
  tau1 ~ dgamma(0.1, 0.1)
  tau2 ~ dgamma(0.1, 0.1)

  # Converting precision to standard deviation
  sigma0 <- 1 / sqrt(tau0)
  sigma1 <- 1 / sqrt(tau1)
  sigma2 <- 1 / sqrt(tau2)
}
"

writeLines(jags_model_seasonal, "jags_model_seasonal.txt")
```


#### Seasonal model for fox
```{r message=FALSE, warning=FALSE}
# This model includes temperature + rainfall + monthly seasonal effects

fox_jags_data_seasonal <- list(
  y        = fox_data$total_sightings,
  temp     = fox_data$temp_scaled,
  rainfall = fox_data$rainfall_scaled,
  month    = fox_data$month,
  site     = fox_data$site,
  N        = nrow(fox_data),
  S        = max(fox_data$site)
)

inits_seasonal <- list(
  beta0 = 0, beta1 = 0,
  beta2 = 0, beta3 = 0,
  tau0  = 1, tau1  = 1, tau2 = 1
)

set.seed(123)

fox_model_seasonal <- jags.model(
  file     = "jags_model_seasonal.txt",
  data     = fox_jags_data_seasonal,
  inits    = inits_seasonal,
  n.chains = 3,
  n.adapt  = 1000
)

update(fox_model_seasonal, n.iter = 5000)

fox_samples_seasonal <- coda.samples(
  model          = fox_model_seasonal,
  variable.names = c("beta0", "beta1", "beta2", "beta3",
                     "alpha", "sigma0", "sigma1", "sigma2"),
  n.iter         = 10000
)

cat("Fox seasonal model completed\n")
cat("Fox seasonal model convergence check:\n")
print(gelman.diag(fox_samples_seasonal[, c("beta0", "beta1",
                                            "beta2", "beta3",
                                            "sigma0", "sigma1",
                                            "sigma2")]))

cat("\nFox seasonal model summary (main parameters):\n")
fox_seasonal_summary <- summary(fox_samples_seasonal)
print(fox_seasonal_summary$statistics[
  c("beta0", "beta1", "beta2", "beta3",
    "sigma0", "sigma1", "sigma2"), ])
```


#### Seasonal model for hare
```{r message=FALSE, warning=FALSE}
hare_jags_data_seasonal <- list(
  y        = hare_data$total_sightings,
  temp     = hare_data$temp_scaled,
  rainfall = hare_data$rainfall_scaled,
  month    = hare_data$month,
  site     = hare_data$site,
  N        = nrow(hare_data),
  S        = max(hare_data$site)
)

set.seed(123)

hare_model_seasonal <- jags.model(
  file     = "jags_model_seasonal.txt",
  data     = hare_jags_data_seasonal,
  inits    = inits_seasonal,
  n.chains = 3,
  n.adapt  = 1000
)

update(hare_model_seasonal, n.iter = 5000)

hare_samples_seasonal <- coda.samples(
  model          = hare_model_seasonal,
  variable.names = c("beta0", "beta1", "beta2", "beta3",
                     "alpha", "sigma0", "sigma1", "sigma2"),
  n.iter         = 10000
)

cat("Hare seasonal model is completed\n")
cat("\nConvergence check:\n")
print(gelman.diag(hare_samples_seasonal[, c("beta0", "beta1",
                                             "beta2", "beta3",
                                             "sigma0", "sigma1",
                                             "sigma2")]))

cat("\nHare main parameters:\n")
hare_seasonal_summary <- summary(hare_samples_seasonal)
print(hare_seasonal_summary$statistics[
  c("beta0", "beta1", "beta2", "beta3",
    "sigma0", "sigma1", "sigma2"), ])
```



#### Seasonal model for pine marten
```{r message=FALSE, warning=FALSE}
marten_jags_data_seasonal <- list(
  y        = marten_data$total_sightings,
  temp     = marten_data$temp_scaled,
  rainfall = marten_data$rainfall_scaled,
  month    = marten_data$month,
  site     = marten_data$site,
  N        = nrow(marten_data),
  S        = max(marten_data$site)
)

set.seed(123)

marten_model_seasonal <- jags.model(
  file     = "jags_model_seasonal.txt",
  data     = marten_jags_data_seasonal,
  inits    = inits_seasonal,
  n.chains = 3,
  n.adapt  = 1000
)

update(marten_model_seasonal, n.iter = 5000)

marten_samples_seasonal <- coda.samples(
  model          = marten_model_seasonal,
  variable.names = c("beta0", "beta1", "beta2", "beta3",
                     "alpha", "sigma0", "sigma1", "sigma2"),
  n.iter         = 10000
)

cat("Pine marten seasonal model is completed\n")
cat("\nConvergence check:\n")
print(gelman.diag(marten_samples_seasonal[, c("beta0", "beta1",
                                               "beta2", "beta3",
                                               "sigma0", "sigma1",
                                               "sigma2")]))

cat("\nMarten main parameters:\n")
marten_seasonal_summary <- summary(marten_samples_seasonal)
print(marten_seasonal_summary$statistics[
  c("beta0", "beta1", "beta2", "beta3",
    "sigma0", "sigma1", "sigma2"), ])
```


#### Extended MCMC for hare seasonal model
```{r message=FALSE, warning=FALSE}
# Hare seasonal model needs more iterations - beta0 R-hat is 1.19
cat("Running extended MCMC for hare seasonal model...\n")

update(hare_model_seasonal, n.iter = 5000)

hare_samples_seasonal2 <- coda.samples(
  model          = hare_model_seasonal,
  variable.names = c("beta0", "beta1", "beta2", "beta3",
                     "alpha", "sigma0", "sigma1", "sigma2"),
  n.iter         = 10000
)

cat("Extended hare seasonal MCMC complete!\n")
cat("\nConvergence check:\n")
print(gelman.diag(hare_samples_seasonal2[, c("beta0", "beta1",
                                              "beta2", "beta3",
                                              "sigma0", "sigma1",
                                              "sigma2")]))
cat("\nHare seasonal main parameters:\n")
hare_seasonal_summary2 <- summary(hare_samples_seasonal2)
print(hare_seasonal_summary2$statistics[
  c("beta0", "beta1", "beta2", "beta3",
    "sigma0", "sigma1", "sigma2"), ])
```


#### Seasonal effects plot
```{r message=FALSE, warning=FALSE, fig.width=14, fig.height=6}
# Extracting the monthly alpha effects for each species
# alpha[1] = 0 (January baseline)
# alpha[2] to alpha[12] show how each month differs from January

# Function to extract monthly effects
get_monthly_effects <- function(samples_summary, species_name) {
  # Getting alpha parameters
  alpha_names <- paste0("alpha[", 1:12, "]")
  alpha_means <- samples_summary$statistics[alpha_names, "Mean"]
  alpha_lower <- samples_summary$quantiles[alpha_names, "2.5%"]
  alpha_upper <- samples_summary$quantiles[alpha_names, "97.5%"]

  data.frame(
    month   = 1:12,
    month_name = factor(month.abb, levels = month.abb),
    effect  = alpha_means,
    lower   = alpha_lower,
    upper   = alpha_upper,
    species = species_name
  )
}

# Getting summaries for all three species
fox_seasonal_sum    <- summary(fox_samples_seasonal)
hare_seasonal_sum   <- summary(hare_samples_seasonal2)
marten_seasonal_sum <- summary(marten_samples_seasonal)

# Extracting monthly effects
fox_monthly    <- get_monthly_effects(fox_seasonal_sum,
                                      "Vulpes vulpes (Fox)")
hare_monthly   <- get_monthly_effects(hare_seasonal_sum,
                                      "Lepus timidus (Hare)")
marten_monthly <- get_monthly_effects(marten_seasonal_sum,
                                      "Martes martes (Pine marten)")

# Combining
all_monthly <- dplyr::bind_rows(fox_monthly, hare_monthly, marten_monthly)

# Plotting monthly effects with credible intervals
ggplot(all_monthly, aes(x = month_name, y = effect,
                         group = species, colour = species)) +
  geom_hline(yintercept = 0, linetype = "dashed",
             colour = "grey50", linewidth = 0.8) +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = species),
              alpha = 0.2) +
  geom_line(linewidth = 1.0) +
  geom_point(size = 2) +
  facet_wrap(~species, ncol = 3) +
  labs(
    title    = "Monthly seasonal effects on sightings - all three species",
    subtitle = "Relative to January baseline (dashed line = no seasonal effect)",
    x        = "Month",
    y        = "Seasonal effect (log scale)"
  ) +
  theme_minimal(base_size = 16) +
  theme(
    legend.position  = "none",
    axis.text.x      = element_text(angle = 45, hjust = 1)
  )

ggsave("plot_seasonal_effects.png", width = 14, height = 6,dpi = 200)

```


#### Saving seasonal model results for all three species
```{r message=FALSE, warning=FALSE}
# Fox
fox_seasonal_results <- data.frame(
  parameter = rownames(fox_seasonal_sum$statistics),
  mean      = fox_seasonal_sum$statistics[, "Mean"],
  sd        = fox_seasonal_sum$statistics[, "SD"],
  lower_95  = fox_seasonal_sum$quantiles[, "2.5%"],
  upper_95  = fox_seasonal_sum$quantiles[, "97.5%"]
)
write.csv(fox_seasonal_results, "fox_seasonal_model_results.csv")

# Hare
hare_seasonal_results <- data.frame(
  parameter = rownames(hare_seasonal_sum$statistics),
  mean      = hare_seasonal_sum$statistics[, "Mean"],
  sd        = hare_seasonal_sum$statistics[, "SD"],
  lower_95  = hare_seasonal_sum$quantiles[, "2.5%"],
  upper_95  = hare_seasonal_sum$quantiles[, "97.5%"]
)
write.csv(hare_seasonal_results, "hare_seasonal_model_results.csv")

# Pine marten
marten_seasonal_results <- data.frame(
  parameter = rownames(marten_seasonal_sum$statistics),
  mean      = marten_seasonal_sum$statistics[, "Mean"],
  sd        = marten_seasonal_sum$statistics[, "SD"],
  lower_95  = marten_seasonal_sum$quantiles[, "2.5%"],
  upper_95  = marten_seasonal_sum$quantiles[, "97.5%"]
)
write.csv(marten_seasonal_results, "marten_seasonal_model_results.csv")

cat("All seasonal model results saved!\n")
cat("\nFiles saved:\n")
cat("- fox_seasonal_model_results.csv\n")
cat("- hare_seasonal_model_results.csv\n")
cat("- marten_seasonal_model_results.csv\n")
```



