---
title: "ST606 Project - 7. Diagnostics"
author: "Kessiya Mary Jose - 25250360"
date: "`r format(Sys.time(), '%X %d %B, %Y')`"
output: html_document
---

#### Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE,
                      cache.path = "cache/")
library(tidyverse)
library(rjags)
library(coda)
library(parallel)
library(patchwork)

options(mc.cores = 6)
```


#### Load dataset and prepare species data
```{r message=FALSE, warning=FALSE}
# Loading the final dataset and preparing the three species datasets

final_data <- read_csv("final_dataset_expanded2.csv",
                       show_col_types = FALSE) %>%
  dplyr::mutate(
    date            = as.Date(paste(year, month, "01", sep = "-")),
    temp_scaled     = as.numeric(scale(mean_temp)),
    rainfall_scaled = as.numeric(scale(rainfall_mm)),
    site            = as.integer(factor(station))
  )

fox_data <- final_data %>%
  dplyr::filter(species == "Vulpes vulpes",
                total_sightings >= 1)

hare_data <- final_data %>%
  dplyr::filter(species == "Lepus timidus subsp. hibernicus",
                total_sightings >= 1)

marten_data <- final_data %>%
  dplyr::filter(species == "Martes martes",
                total_sightings >= 1)

cat("Fox rows:", nrow(fox_data), "\n")
cat("Hare rows:", nrow(hare_data), "\n")
cat("Marten rows:", nrow(marten_data), "\n")
```


#### Refit combined model for diagnostics
```{r message=FALSE, warning=FALSE}
# Refitting the combined model for all three species
# We need the model objects in memory for diagnostic checks

inits_combined <- list(
  beta0 = 0, beta1 = 0,
  beta2 = 0, beta3 = 0,
  tau0  = 1, tau1  = 1, tau2 = 1
)

# Fox
fox_diag_data <- list(
  y        = fox_data$total_sightings,
  temp     = fox_data$temp_scaled,
  rainfall = fox_data$rainfall_scaled,
  month    = fox_data$month,
  site     = fox_data$site,
  N        = nrow(fox_data),
  S        = max(fox_data$site)
)

set.seed(123)
fox_diag_model <- jags.model(
  file     = "jags_model_combined.txt",
  data     = fox_diag_data,
  inits    = inits_combined,
  n.chains = 3,
  n.adapt  = 1000
)
update(fox_diag_model, n.iter = 5000)
fox_diag_samples <- coda.samples(
  model          = fox_diag_model,
  variable.names = c("beta0", "beta1", "beta2", "beta3",
                     "alpha", "sigma0", "sigma1", "sigma2"),
  n.iter         = 10000,
  thin           = 2
)
cat("Fox model ready for diagnostics\n")

# Hare
hare_diag_data <- list(
  y        = hare_data$total_sightings,
  temp     = hare_data$temp_scaled,
  rainfall = hare_data$rainfall_scaled,
  month    = hare_data$month,
  site     = hare_data$site,
  N        = nrow(hare_data),
  S        = max(hare_data$site)
)

set.seed(123)
hare_diag_model <- jags.model(
  file     = "jags_model_combined.txt",
  data     = hare_diag_data,
  inits    = inits_combined,
  n.chains = 3,
  n.adapt  = 1000
)
update(hare_diag_model, n.iter = 8000)
hare_diag_samples <- coda.samples(
  model          = hare_diag_model,
  variable.names = c("beta0", "beta1", "beta2", "beta3",
                     "alpha", "sigma0", "sigma1", "sigma2"),
  n.iter         = 15000,
  thin           = 2
)
cat("Hare model ready for diagnostics\n")

# Pine marten
marten_diag_data <- list(
  y        = marten_data$total_sightings,
  temp     = marten_data$temp_scaled,
  rainfall = marten_data$rainfall_scaled,
  month    = marten_data$month,
  site     = marten_data$site,
  N        = nrow(marten_data),
  S        = max(marten_data$site)
)

set.seed(123)
marten_diag_model <- jags.model(
  file     = "jags_model_combined.txt",
  data     = marten_diag_data,
  inits    = inits_combined,
  n.chains = 3,
  n.adapt  = 1000
)
update(marten_diag_model, n.iter = 5000)
marten_diag_samples <- coda.samples(
  model          = marten_diag_model,
  variable.names = c("beta0", "beta1", "beta2", "beta3",
                     "alpha", "sigma0", "sigma1", "sigma2"),
  n.iter         = 10000,
  thin           = 2
)
cat("Pine marten model ready for diagnostics\n")
```


#### Prior-posterior overlap check
```{r message=FALSE, warning=FALSE, fig.width=14, fig.height=8}
# This checks whether the flat priors are influencing the posterior
# If the posterior is very different from the prior, the data is driving the estimates 
# If they overlap completely, the prior is too informative

# I used flat normal priors: dnorm(0, 0.001)
# which means mean=0, precision=0.001, so SD = sqrt(1/0.001) = 31.6
# This is a very wide prior - almost flat

# Generating prior samples for comparison
set.seed(123)
n_samples <- 15000
prior_samples <- rnorm(n_samples, mean = 0, sd = sqrt(1/0.001))

# Extracting posterior samples for beta0, beta1, beta2, beta3
# for each species
get_posterior <- function(samples, param) {
  as.numeric(as.matrix(samples)[, param])
}

# Creating data for plotting
prior_df <- data.frame(
  value   = prior_samples,
  type    = "Prior",
  species = "All species"
)

# Fox posteriors
fox_post <- data.frame(
  rbind(
    data.frame(value = get_posterior(fox_diag_samples, "beta0"),
               parameter = "Intercept (β₀)"),
    data.frame(value = get_posterior(fox_diag_samples, "beta1"),
               parameter = "Temperature linear (β₁)"),
    data.frame(value = get_posterior(fox_diag_samples, "beta2"),
               parameter = "Temperature quadratic (β₂)"),
    data.frame(value = get_posterior(fox_diag_samples, "beta3"),
               parameter = "Rainfall (β₃)")
  ),
  species = "Vulpes vulpes (Fox)"
)

# Hare posteriors
hare_post <- data.frame(
  rbind(
    data.frame(value = get_posterior(hare_diag_samples, "beta0"),
               parameter = "Intercept (β₀)"),
    data.frame(value = get_posterior(hare_diag_samples, "beta1"),
               parameter = "Temperature linear (β₁)"),
    data.frame(value = get_posterior(hare_diag_samples, "beta2"),
               parameter = "Temperature quadratic (β₂)"),
    data.frame(value = get_posterior(hare_diag_samples, "beta3"),
               parameter = "Rainfall (β₃)")
  ),
  species = "Lepus timidus (Hare)"
)

# Marten posteriors
marten_post <- data.frame(
  rbind(
    data.frame(value = get_posterior(marten_diag_samples, "beta0"),
               parameter = "Intercept (β₀)"),
    data.frame(value = get_posterior(marten_diag_samples, "beta1"),
               parameter = "Temperature linear (β₁)"),
    data.frame(value = get_posterior(marten_diag_samples, "beta2"),
               parameter = "Temperature quadratic (β₂)"),
    data.frame(value = get_posterior(marten_diag_samples, "beta3"),
               parameter = "Rainfall (β₃)")
  ),
  species = "Martes martes (Pine marten)"
)

# Combining all posteriors
all_post <- dplyr::bind_rows(fox_post, hare_post, marten_post)

# Plotting prior vs posterior for each parameter and species
ggplot(all_post, aes(x = value, fill = species, colour = species)) +
  # Prior shown as dashed line
  geom_density(data = data.frame(
    value = prior_samples,
    parameter = rep(c("Intercept (β₀)",
                      "Temperature linear (β₁)",
                      "Temperature quadratic (β₂)",
                      "Rainfall (β₃)"),
                    each = n_samples/4)),
    aes(x = value), fill = NA,
    colour = "black", linetype = "dashed",
    linewidth = 0.8, inherit.aes = FALSE) +
  # Posteriors shown as filled densities
  geom_density(alpha = 0.3, linewidth = 0.8) +
  facet_grid(species ~ parameter, scales = "free") +
  coord_cartesian(xlim = c(-3, 3)) +
  labs(
    title    = "Prior-posterior overlap check",
    subtitle = "Dashed line = prior distribution, filled = posterior. Good model: posterior clearly different from prior",
    x        = "Parameter value",
    y        = "Density"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

ggsave("plot_prior_posterior_overlap.png", width = 14, height = 8)
cat("Saved: plot_prior_posterior_overlap.png\n")
```


#### Posterior predictive check
```{r message=FALSE, warning=FALSE, fig.width=14, fig.height=6}
# Function to simulate full datasets from the model
run_ppc_density <- function(diag_model, observed_y,
                             species_name, n_sim = 100) {
  # Sampling mu values from the model
  ppc_samples <- coda.samples(
    model          = diag_model,
    variable.names = "mu",
    n.iter         = n_sim,
    thin           = 1
  )

  mu_matrix <- as.matrix(ppc_samples)

  # Simulating full datasets for each set of mu values
  sim_datasets <- apply(mu_matrix, 1, function(mu_row) {
    y_sim <- rpois(length(mu_row), lambda = mu_row)
    y_sim[y_sim < 1] <- 1
    y_sim
  })

  list(
    observed     = observed_y,
    sim_datasets = sim_datasets,
    species      = species_name
  )
}

cat("Running PPC density for fox...\n")
fox_ppc_d    <- run_ppc_density(fox_diag_model,
                                 fox_data$total_sightings,
                                 "Vulpes vulpes (Fox)")

cat("Running PPC density for hare...\n")
hare_ppc_d   <- run_ppc_density(hare_diag_model,
                                  hare_data$total_sightings,
                                  "Lepus timidus (Hare)")

cat("Running PPC density for pine marten...\n")
marten_ppc_d <- run_ppc_density(marten_diag_model,
                                  marten_data$total_sightings,
                                  "Martes martes (Pine marten)")

# Function to create density plot for one species
plot_ppc_density <- function(ppc_result, colour) {
  obs     <- ppc_result$observed
  sim_mat <- ppc_result$sim_datasets
  species <- ppc_result$species

  # Using only 50 simulated lines so they are visible
  n_show <- min(50, ncol(sim_mat))
  idx    <- sample(1:ncol(sim_mat), n_show)

  sim_df <- do.call(rbind, lapply(idx, function(i) {
    d <- density(sim_mat[, i], from = 1, to = max(obs) + 2)
    data.frame(x = d$x, y = d$y, sim = i)
  }))

  obs_d  <- density(obs, from = 1, to = max(obs) + 2)
  obs_df <- data.frame(x = obs_d$x, y = obs_d$y)

  ggplot() +
    geom_line(data = sim_df,
              aes(x = x, y = y, group = sim),
              colour = colour, alpha = 0.25,
              linewidth = 0.6) +
    geom_line(data = obs_df,
              aes(x = x, y = y),
              colour = "black",
              linewidth = 1.5) +
    labs(title   = species,
         x       = "Sightings",
         y       = "Density") +
    theme_minimal() +
    theme(plot.title = element_text(face = "italic"))
}

p_fox    <- plot_ppc_density(fox_ppc_d,    "#2196F3")
p_hare   <- plot_ppc_density(hare_ppc_d,   "#E53935")
p_marten <- plot_ppc_density(marten_ppc_d, "#43A047")

combined_ppc <- p_fox + p_hare + p_marten +
  plot_annotation(
    title    = "Posterior predictive check - density plot",
    subtitle = "Black line = observed data. Coloured lines = simulated datasets from model.\nGood fit: black line sits within the bundle of coloured lines",
    theme    = theme(
      plot.title    = element_text(size = 14, face = "bold"),
      plot.subtitle = element_text(size = 11)
    )
  )

print(combined_ppc)
ggsave("plot_ppc_density.png", combined_ppc, width = 14, height = 5)
cat("Saved: plot_ppc_density.png\n")
```


#### Log-scale posterior predictive check
```{r message=FALSE, warning=FALSE, fig.width=14, fig.height=6}
# Redoing the PPC on log scale to better visualise the distribution shape
# Log transformation spreads out the right-skewed sighting data making the density plots look more like the typical bell shape
# This makes it easier to see whether the model captures the data distribution

plot_ppc_density_log <- function(ppc_result, colour) {
  obs     <- ppc_result$observed
  sim_mat <- ppc_result$sim_datasets
  species <- ppc_result$species

  n_show <- min(50, ncol(sim_mat))
  idx    <- sample(1:ncol(sim_mat), n_show)

  sim_df <- do.call(rbind, lapply(idx, function(i) {
    d <- density(log(sim_mat[, i] + 0.5))
    data.frame(x = d$x, y = d$y, sim = i)
  }))

  obs_d  <- density(log(obs + 0.5))
  obs_df <- data.frame(x = obs_d$x, y = obs_d$y)

  ggplot() +
    geom_line(data = sim_df,
              aes(x = x, y = y, group = sim),
              colour = colour, alpha = 0.3,
              linewidth = 0.5) +
    geom_line(data = obs_df,
              aes(x = x, y = y),
              colour = "black",
              linewidth = 1.5) +
    labs(title = species,
         x     = "log(Sightings)",
         y     = "Density") +
    theme_minimal() +
    theme(plot.title = element_text(face = "italic"))
}

p_fox_log    <- plot_ppc_density_log(fox_ppc_d,    "#2196F3")
p_hare_log   <- plot_ppc_density_log(hare_ppc_d,   "#E53935")
p_marten_log <- plot_ppc_density_log(marten_ppc_d, "#43A047")

combined_ppc_log <- p_fox_log + p_hare_log + p_marten_log +
  plot_annotation(
    title    = "Posterior predictive check - log scale density plot",
    subtitle = "Black line = observed data. Coloured lines = simulated datasets from model.\n Good fit: black line sits within the bundle of coloured lines",
    theme    = theme(
      plot.title    = element_text(size = 14, face = "bold"),
      plot.subtitle = element_text(size = 11)
    )
  )

print(combined_ppc_log)
ggsave("plot_ppc_density_log.png", combined_ppc_log, width = 14, height = 5)
cat("Saved: plot_ppc_density_log.png\n")
```


#### Fitted versus observed plot
```{r message=FALSE, warning=FALSE, fig.width=14, fig.height=5}
# shows how well the model predicted each individual observation
# Points close to the diagonal line = good fit
# Points far from diagonal = poor fit for those observations

# Function to get fitted values for one species
get_fitted <- function(diag_samples, species_data, species_name) {

  # Getting posterior mean of mu for each observation
  samples_matrix <- as.matrix(diag_samples)

  # Finding columns that start with "mu["
  mu_cols <- grep("^mu\\[", colnames(samples_matrix), value = TRUE)

  if (length(mu_cols) == 0) {
    cat("No mu columns found - need to refit with mu in variable.names\n")
    return(NULL)
  }

  mu_means <- colMeans(samples_matrix[, mu_cols])

  data.frame(
    observed = species_data$total_sightings,
    fitted   = mu_means,
    species  = species_name
  )
}

# need to refit sampling mu explicitly
cat("Extracting fitted values for fox...\n")
fox_mu_samples <- coda.samples(
  model          = fox_diag_model,
  variable.names = "mu",
  n.iter         = 2000,
  thin           = 1
)

cat("Extracting fitted values for hare...\n")
hare_mu_samples <- coda.samples(
  model          = hare_diag_model,
  variable.names = "mu",
  n.iter         = 2000,
  thin           = 1
)

cat("Extracting fitted values for pine marten...\n")
marten_mu_samples <- coda.samples(
  model          = marten_diag_model,
  variable.names = "mu",
  n.iter         = 2000,
  thin           = 1
)

# Getting posterior means of mu
fox_fitted <- data.frame(
  observed = fox_data$total_sightings,
  fitted   = exp(colMeans(log(as.matrix(fox_mu_samples)))),
  species  = "Vulpes vulpes (Fox)"
)

hare_fitted <- data.frame(
  observed = hare_data$total_sightings,
  fitted   = exp(colMeans(log(as.matrix(hare_mu_samples)))),
  species  = "Lepus timidus (Hare)"
)

marten_fitted <- data.frame(
  observed = marten_data$total_sightings,
  fitted   = exp(colMeans(log(as.matrix(marten_mu_samples)))),
  species  = "Martes martes (Pine marten)"
)

all_fitted <- dplyr::bind_rows(fox_fitted, hare_fitted, marten_fitted)

# Plotting fitted vs observed
ggplot(all_fitted, aes(x = fitted, y = observed)) +
  geom_point(alpha = 0.3, colour = "#1565C0", size = 1.5) +
  geom_abline(intercept = 0, slope = 1,
              colour = "#E53935", linewidth = 1.0,
              linetype = "dashed") +
  facet_wrap(~species, scales = "free", ncol = 3) +
  labs(
    title    = "Fitted versus observed sightings",
    subtitle = "Red dashed line = perfect fit. Points close to line = good prediction",
    x        = "Fitted values (posterior mean of μ)",
    y        = "Observed sightings"
  ) +
  theme_minimal()

ggsave("plot_fitted_vs_observed.png", width = 14, height = 5)
cat("Saved: plot_fitted_vs_observed.png\n")
```


#### Truncated negative binomial model - setup
```{r message=FALSE, warning=FALSE}
# Trying a truncated negative binomial model because the posterior predictive checks showed mild over-dispersion

# The JAGS model code for truncated negative binomial
jags_negbin_code <- "
model {
  for (i in 1:N) {
    # Truncated negative binomial - minimum value is 1
    y[i] ~ dnegbin(p[i], r) T(1,)
    # Converting to the mean parameterisation
    # mu = r * (1 - p) / p
    # so p = r / (r + mu)
    p[i] <- r / (r + mu[i])
    log(mu[i]) <- (beta0 + b0[site[i]]) +
                  (beta1 + b1[site[i]]) * temp[i] +
                  (beta2 + b2[site[i]]) * temp[i]^2 +
                  beta3 * rainfall[i] +
                  alpha[month[i]]
  }

  # Random effects for each site
  for (s in 1:S) {
    b0[s] ~ dnorm(0, tau0)
    b1[s] ~ dnorm(0, tau1)
    b2[s] ~ dnorm(0, tau2)
  }

  # Monthly seasonal effects
  alpha[1] <- 0
  for (m in 2:12) {
    alpha[m] ~ dnorm(0, 0.001)
  }

  # Fixed effect priors
  beta0 ~ dnorm(0, 0.001)
  beta1 ~ dnorm(0, 0.001)
  beta2 ~ dnorm(0, 0.001)
  beta3 ~ dnorm(0, 0.001)

  # Overdispersion parameter
  # r > 0, smaller r means more overdispersion
  # r -> infinity means Poisson (no overdispersion)
  r ~ dgamma(0.1, 0.1)

  # Variance components
  tau0 ~ dgamma(0.1, 0.1)
  tau1 ~ dgamma(0.1, 0.1)
  tau2 ~ dgamma(0.1, 0.1)

  sigma0 <- 1 / sqrt(tau0)
  sigma1 <- 1 / sqrt(tau1)
  sigma2 <- 1 / sqrt(tau2)
}
"

writeLines(jags_negbin_code, "jags_model_negbin.txt")
cat("Negative binomial model code saved to jags_model_negbin.txt\n")
```


#### Truncated negative binomial - fox
```{r message=FALSE, warning=FALSE}
# Fitting the truncated negative binomial for fox

inits_negbin <- list(
  beta0 = 0, beta1 = 0,
  beta2 = 0, beta3 = 0,
  tau0  = 1, tau1  = 1, tau2 = 1,
  r     = 1
)

fox_negbin_data <- list(
  y        = fox_data$total_sightings,
  temp     = fox_data$temp_scaled,
  rainfall = fox_data$rainfall_scaled,
  month    = fox_data$month,
  site     = fox_data$site,
  N        = nrow(fox_data),
  S        = max(fox_data$site)
)

set.seed(123)

fox_negbin_model <- jags.model(
  file     = "jags_model_negbin.txt",
  data     = fox_negbin_data,
  inits    = inits_negbin,
  n.chains = 3,
  n.adapt  = 1000
)

update(fox_negbin_model, n.iter = 5000)

fox_negbin_samples <- coda.samples(
  model          = fox_negbin_model,
  variable.names = c("beta0", "beta1", "beta2", "beta3",
                     "r", "sigma0", "sigma1", "sigma2"),
  n.iter         = 10000,
  thin           = 2
)

cat("Fox negative binomial model complete!\n")
cat("\nConvergence check:\n")
print(gelman.diag(fox_negbin_samples))
cat("\nResults:\n")
print(summary(fox_negbin_samples))
```


#### Truncated negative binomial - hare
```{r message=FALSE, warning=FALSE}
# Fitting the truncated negative binomial for hare

hare_negbin_data <- list(
  y        = hare_data$total_sightings,
  temp     = hare_data$temp_scaled,
  rainfall = hare_data$rainfall_scaled,
  month    = hare_data$month,
  site     = hare_data$site,
  N        = nrow(hare_data),
  S        = max(hare_data$site)
)

set.seed(123)

hare_negbin_model <- jags.model(
  file     = "jags_model_negbin.txt",
  data     = hare_negbin_data,
  inits    = inits_negbin,
  n.chains = 3,
  n.adapt  = 1000
)

update(hare_negbin_model, n.iter = 8000)

hare_negbin_samples <- coda.samples(
  model          = hare_negbin_model,
  variable.names = c("beta0", "beta1", "beta2", "beta3",
                     "r", "sigma0", "sigma1", "sigma2"),
  n.iter         = 15000,
  thin           = 2
)

cat("Hare negative binomial model complete!\n")
cat("\nConvergence check:\n")
print(gelman.diag(hare_negbin_samples))
cat("\nResults:\n")
print(summary(hare_negbin_samples))
```


#### Truncated negative binomial - pine marten
```{r message=FALSE, warning=FALSE}
# Fitting the truncated negative binomial for pine marten

marten_negbin_data <- list(
  y        = marten_data$total_sightings,
  temp     = marten_data$temp_scaled,
  rainfall = marten_data$rainfall_scaled,
  month    = marten_data$month,
  site     = marten_data$site,
  N        = nrow(marten_data),
  S        = max(marten_data$site)
)

set.seed(123)

marten_negbin_model <- jags.model(
  file     = "jags_model_negbin.txt",
  data     = marten_negbin_data,
  inits    = inits_negbin,
  n.chains = 3,
  n.adapt  = 1000
)

update(marten_negbin_model, n.iter = 5000)

marten_negbin_samples <- coda.samples(
  model          = marten_negbin_model,
  variable.names = c("beta0", "beta1", "beta2", "beta3",
                     "r", "sigma0", "sigma1", "sigma2"),
  n.iter         = 10000,
  thin           = 2
)

cat("Pine marten negative binomial model complete!\n")
cat("\nConvergence check:\n")
print(gelman.diag(marten_negbin_samples))
cat("\nResults:\n")
print(summary(marten_negbin_samples))
```


#### Negative binomial - to save results and compare with Poisson
```{r message=FALSE, warning=FALSE}
# Saving negative binomial results
fox_negbin_sum    <- summary(fox_negbin_samples)
hare_negbin_sum   <- summary(hare_negbin_samples)
marten_negbin_sum <- summary(marten_negbin_samples)

# Loading the Poisson results from the saved CSV file
poisson_results <- read_csv("combined_model_coefficients.csv",
                             show_col_types = FALSE)

params_to_compare <- c("beta0", "beta1", "beta2", "beta3")

# Extracting negbin results
negbin_results <- dplyr::bind_rows(
  data.frame(
    species       = "Vulpes vulpes (Fox)",
    parameter     = params_to_compare,
    negbin_mean   = round(fox_negbin_sum$statistics[params_to_compare, "Mean"], 3),
    negbin_lower  = round(fox_negbin_sum$quantiles[params_to_compare, "2.5%"], 3),
    negbin_upper  = round(fox_negbin_sum$quantiles[params_to_compare, "97.5%"], 3)
  ),
  data.frame(
    species       = "Lepus timidus (Hare)",
    parameter     = params_to_compare,
    negbin_mean   = round(hare_negbin_sum$statistics[params_to_compare, "Mean"], 3),
    negbin_lower  = round(hare_negbin_sum$quantiles[params_to_compare, "2.5%"], 3),
    negbin_upper  = round(hare_negbin_sum$quantiles[params_to_compare, "97.5%"], 3)
  ),
  data.frame(
    species       = "Martes martes (Pine marten)",
    parameter     = params_to_compare,
    negbin_mean   = round(marten_negbin_sum$statistics[params_to_compare, "Mean"], 3),
    negbin_lower  = round(marten_negbin_sum$quantiles[params_to_compare, "2.5%"], 3),
    negbin_upper  = round(marten_negbin_sum$quantiles[params_to_compare, "97.5%"], 3)
  )
)

# Joining with Poisson results
poisson_clean <- poisson_results %>%
  dplyr::filter(parameter %in% params_to_compare) %>%
  dplyr::select(species, parameter,
                poisson_mean  = posterior_mean,
                poisson_lower = lower_95,
                poisson_upper = upper_95)

comparison_table <- poisson_clean %>%
  dplyr::left_join(negbin_results,
                   by = c("species", "parameter"))

write_csv(comparison_table, "poisson_vs_negbin_comparison.csv")
cat("Saved: poisson_vs_negbin_comparison.csv\n")

cat("\nComparison table:\n")
print(comparison_table)

cat("\nOverdispersion r summary:\n")
r_summary <- data.frame(
  species   = c("Vulpes vulpes (Fox)",
                "Lepus timidus (Hare)",
                "Martes martes (Pine marten)"),
  r_mean    = round(c(
    fox_negbin_sum$statistics["r", "Mean"],
    hare_negbin_sum$statistics["r", "Mean"],
    marten_negbin_sum$statistics["r", "Mean"]), 3),
  r_lower   = round(c(
    fox_negbin_sum$quantiles["r", "2.5%"],
    hare_negbin_sum$quantiles["r", "2.5%"],
    marten_negbin_sum$quantiles["r", "2.5%"]), 3),
  r_upper   = round(c(
    fox_negbin_sum$quantiles["r", "97.5%"],
    hare_negbin_sum$quantiles["r", "97.5%"],
    marten_negbin_sum$quantiles["r", "97.5%"]), 3)
)
print(r_summary)
write_csv(r_summary, "negbin_r_summary.csv")
cat("Saved: negbin_r_summary.csv\n")
```



